---
title: "Different approaches to quantify years of life lost from COVID-19"
author: "Tamás Ferenci"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 12, dpi = 300)
options(digits = 3, knitr.kable.NA = "", scipen = 999)
```

## Manuscript:

Preprint is available at [https://www.medrxiv.org/content/10.1101/2021.05.13.21257193v1](https://www.medrxiv.org/content/10.1101/2021.05.13.21257193v1).

## Analysis

We first do some preparation:

```{r}
library(ggplot2)
theme_set(theme_bw())
library(data.table)

name_cols <- c("atr_fib", "cancer", "copd", "demen", "diab", "h_fail", "hypert", "ihd", "renal", "liver",
               "stroke")
name_cols_age <- c("atr_fib:age", "cancer:age", "copd:age", "demen:age", "diab:age", "h_fail:age", 
                   "hypert:age", "ihd:age", "renal:age", "liver:age", "stroke:age")

StartDate <- as.Date("2020-03-04")
EndDate <- Sys.Date()
Pop50plus <- 3828819
```

We download the life table:

```{r}
if(!file.exists("LifeTableHUN.rds")) {
  LifeTableMale <- HMDHFDplus::readHMDweb("HUN", "mltper_1x1")
  LifeTableFemale <- HMDHFDplus::readHMDweb("HUN", "fltper_1x1")
  LifeTable <- rbind(cbind(Sex = "Male", LifeTableMale), cbind(Sex = "Female", LifeTableFemale))
  LifeTable <- data.table(LifeTable[LifeTable$Year==2017, ])
  saveRDS(LifeTable, "LifeTableHUN.rds")
} else LifeTable <- readRDS("LifeTableHUN.rds")
```

We then scrape the mortality data:

```{r}
pg <- xml2::read_html("https://koronavirus.gov.hu/elhunytak")
lastpg <- strsplit(rvest::html_attr(rvest::html_node(pg, xpath = "//li[@class='pager-last']/a"), "href"),
                   "=")[[1]][2]
MortData <- rbindlist(lapply(0:lastpg, function(i)
  rvest::html_table(xml2::read_html(paste0("https://koronavirus.gov.hu/elhunytak?page=", i)))[[1]]))
names(MortData) <- c("ID", "Sex", "Age", "Comorbidities")
unique(MortData$Sex)
MortData$Sex <- as.factor(ifelse(MortData$Sex%in%c("férfi", "Férfi"), "Male", "Female"))
saveRDS(MortData, "MortData.rds")
```

Basic data:

```{r}
nrow(MortData)
MortData <- MortData[Age>=50]
table(MortData$Sex)
nrow(MortData)
```

We create the first plot and save it:

```{r}
ggplot(MortData, aes(x = Age, color = Sex)) + stat_density(geom = "line", position = "identity") +
  labs(x = "Age [year]", y = "") +
  theme(legend.position = "bottom", legend.title = element_blank())
ggsave("Fig1.eps", width = 84, height = 84, units = "mm", device = cairo_ps)
```

We can now merge the life table and the mortality data:

```{r}
MortData <- merge(MortData, LifeTable)
MortData$Sex <- as.factor(MortData$Sex)
```


YLL using the ordinary calculation:

```{r}
sum(MortData$ex) # YLL
tapply(MortData$ex, MortData$Sex, sum)
sum(MortData$ex) / nrow(MortData) # YLL per death
tapply(MortData$ex, MortData$Sex, mean)
sum(MortData$ex) / (as.numeric(difftime(EndDate, StartDate, units = "days"))/365.24 * Pop50plus) # YLL per person-year

```

We then extract the comorbidities:

```{r}
sort(unique(do.call(c, sapply(MortData$Comorbidities, strsplit, split = ""))))

MortData$Comorbidities[grep("\u0096", MortData$Comorbidities)] <- gsub("\u0096", "", MortData$Comorbidities[grep("\u0096", MortData$Comorbidities)])

MortData$Comorbidities <- stringi::stri_trans_general(MortData$Comorbidities, "Latin-ASCII")

sort(unique(do.call(c, sapply(MortData$Comorbidities, strsplit, split = ""))))

MortData$atr_fib <- grepl("fibril", MortData$Comorbidities, ignore.case = TRUE)
MortData$cancer <- grepl("rak|tumor|noma|daganat|metas[z]?t|attet", MortData$Comorbidities, ignore.case = TRUE)
MortData$copd <- grepl("obstru[c|k]tiv[[:blank:]]tudo|copd", MortData$Comorbidities, ignore.case = TRUE)
MortData$demen <- grepl("demen", MortData$Comorbidities, ignore.case = TRUE)
MortData$diab <- grepl("diabetes|cukorbeteg|mellitus", MortData$Comorbidities, ignore.case = TRUE)|grepl("DM|IDDM|T1DM|T2DM", MortData$Comorbidities, ignore.case = FALSE)
MortData$h_fail <- grepl("szivelegtelen", MortData$Comorbidities, ignore.case = TRUE)
MortData$hypert <- grepl("h[i|y]pertonia|magas[[:blank:]|-]?vernyomas", MortData$Comorbidities, ignore.case = TRUE)
MortData$ihd <- grepl("isch[a]?emias[[:blank:]]sziv|iszk[a]?emias[[:blank:]]sziv|angina|infar[k|c]tus", MortData$Comorbidities, ignore.case = TRUE)|grepl("IHD|CAD|CHD|AMI", MortData$Comorbidities, ignore.case = FALSE)
MortData$renal <- grepl("kronikus[[:blank:]]vese[[:blank:]|-]?elegtelen|chronicus[[:blank:]]vese[[:blank:]|-]?elegtelen", MortData$Comorbidities, ignore.case = TRUE)
MortData$liver <- grepl("kronikus[[:blank:]]maj|chronicus[[:blank:]]maj|majzsugor|majelegtelen|cirr|kronikus[[:blank:]]hepa|chronicus[[:blank:]]hepa|majmegna|alkoholos[[:blank:]]maj", MortData$Comorbidities, ignore.case = TRUE)
MortData$stroke <- grepl("stroke|sztrok|agyi[[:blank:]]erkat|agyverzes", MortData$Comorbidities, ignore.case = TRUE)

ComorbLabels <- data.table(variable = c("atr_fib", "cancer", "copd", "demen", "diab",
                                        "h_fail", "hypert", "ihd", "renal", "liver", "stroke"),
                           varlabel = c("Atrial fibrillation", "Cancer", "COPD", "Dementia", "Diabetes", "Heart failure",
                                        "Hypertension", "IHD", "Chronic renal failure", "Chronic liver disease", "Stroke"))
```

We calculate the first table and save it:

```{r}
tab1 <- merge(melt(MortData,  measure.vars = ComorbLabels$variable), ComorbLabels)[
  ,.(mean(value)*100, binom.test(sum(value), length(value))$conf.int[1]*100,
     binom.test(sum(value), length(value))$conf.int[2]*100), .(varlabel)
][, .(varlabel, paste0(round(V1, 1), " (", round(V2, 1), "-", round(V3, 1), ")"))]
knitr::kable(tab1)
write.csv2(tab1, "Tab1.csv")
```

We then create the second figure and save it:

```{r}
predgrid <- expand.grid(Age = seq(50, 100, 0.1), Sex = c("Male", "Female"))
ComorbPrevs <- rbindlist(setNames(lapply(ComorbLabels$variable, function(v) {
  fit <- mgcv::gam(as.formula(paste0(v, "~ s(Age, by = Sex) + Sex")), data = MortData, family = binomial)
  cbind(predgrid, with(predict(fit, predgrid, se.fit = TRUE),
                       data.frame(est = plogis(fit), lwr = plogis(fit - 1.96*se.fit),
                                  upr = plogis(fit + 1.96*se.fit))))
}), ComorbLabels$varlabel), idcol = TRUE)
ggplot(ComorbPrevs, aes(x = Age, y = est*100, color = Sex, fill = Sex, ymin = lwr*100, ymax = upr*100)) +
  geom_line() + geom_ribbon(alpha = 0.2, linetype = 0) + facet_wrap(~.id) +
  labs(x = "Age [year]", y = "Prevalence [%]") +
  theme(legend.position = "bottom", legend.title = element_blank())
ggsave("Fig2.eps", width = 174, height = 174, units = "mm", device = cairo_ps)
```

And then the third one:

```{r}
MortData$ComorbCount <- rowSums(MortData[, ..name_cols])
prop.table(table(MortData$ComorbCount))*100
mean(MortData$ComorbCount>2)*100

ggplot(reshape2::melt(table(MortData$ComorbCount, MortData$Age, MortData$Sex)),
       aes(x = Var2, fill = forcats::fct_rev(as.factor(Var1)), y = value)) + geom_area(position = "fill", alpha = 0.7) +
  facet_grid(cols = vars(Var3)) + scale_x_continuous(expand = c(0, 0), limits = c(50, 90)) +
  scale_y_continuous(labels = scales::percent_format(suffix = ""), expand = c(0, 0)) + 
  labs(x = "Age [year]", y = "Proportion [%]", fill = "Number of comorbidities") +
  guides(fill = guide_legend(reverse = TRUE)) +
  theme(legend.position = "bottom", legend.title = element_blank(), panel.spacing = unit(1, "lines"))
ggsave("Fig3.eps", width = 174, height = 174, units = "mm", device = cairo_ps)
```

Next, we integrate the external survival model and run a few checks:

```{r}
male_coef <- read.csv("https://raw.githubusercontent.com/dmcalli2/covid19_yll_final/master/Data/sail_outputs/male_coef.csv")
male_coef <- setNames(male_coef$x, male_coef$X)
female_coef <- read.csv("https://raw.githubusercontent.com/dmcalli2/covid19_yll_final/master/Data/sail_outputs/female_coef.csv")
female_coef <- setNames(female_coef$x, female_coef$X)

SurvGomp <- function(t, start, ...) {
  ret <- (1 - flexsurv:::pgompertz_work(t, ...))/(1-flexsurv:::pgompertz_work(start, ...))
  ret[t<start] <- 1
  ret
}

times_vect <- seq(50, 150, 0.1)

temp <- 1-SurvGomp(t = times_vect, start = MortData$Age[2], shape = male_coef["shape"],
                   rate = exp(male_coef["rate"] + as.matrix(MortData[2, ..name_cols])%*%male_coef[name_cols] +
                                (MortData$Age[2]*as.matrix(MortData[2, ..name_cols]))%*%male_coef[name_cols_age]),
                   lower_tail = TRUE, give_log = FALSE)

temp <- 1-SurvGomp(t = times_vect, start = 60, shape = female_coef["shape"],
                   rate = exp(female_coef["rate"]),
                   lower_tail = TRUE, give_log = FALSE)

sum(temp - c(0, temp[-length(temp)]))
sum((temp - c(0, temp[-length(temp)]))*times_vect)
sum((temp - c(0, temp[-length(temp)]))*times_vect)-MortData$Age[2]

offsetgrid <- data.table(expand.grid(Sex = c("Male", "Female"), Age = 50:110))
offsetgrid$AdjustedEx <- sapply(1:nrow(offsetgrid), function(i) {
  usedcoefs <- if(offsetgrid$Sex[i]=="Male") male_coef else female_coef
  temp <- 1-SurvGomp(t = times_vect, start = offsetgrid$Age[i], shape = usedcoefs["shape"],
                     rate = exp(usedcoefs["rate"]), lower_tail = TRUE, give_log = FALSE)
  sum((temp - c(0, temp[-length(temp)]))*times_vect)-offsetgrid$Age[i]
})
offsetgrid <- merge(offsetgrid, LifeTable)
ggplot(offsetgrid, aes(x = ex, y = AdjustedEx)) + geom_point() + facet_grid(cols = vars(Sex)) +
  geom_abline(intercept = 0, slope = 1)

offsetgrid$ratio <- offsetgrid$ex/offsetgrid$AdjustedEx
offsetgrid$leratio <- (offsetgrid$Age + offsetgrid$ex)/(offsetgrid$Age + offsetgrid$AdjustedEx)

MortData$AdjustedEx <- sapply(1:nrow(MortData), function(i) {
  usedcoefs <- if(MortData$Sex[i]=="Male") male_coef else female_coef
  temp <- 1-SurvGomp(t = times_vect, start = MortData$Age[i], shape = usedcoefs["shape"],
                     rate = exp(usedcoefs["rate"] + as.matrix(MortData[i, ..name_cols])%*%usedcoefs[name_cols] +
                                  (MortData$Age[i]*as.matrix(MortData[i, ..name_cols]))%*%usedcoefs[name_cols_age]),
                     lower_tail = TRUE, give_log = FALSE)
  # sum((temp - c(0, temp[-length(temp)]))*times_vect)*offsetgrid[Sex==MortData$Sex[i]&Age==MortData$Age[i]]$leratio-MortData$Age[i]
  (sum((temp - c(0, temp[-length(temp)]))*times_vect)-MortData$Age[i])*offsetgrid[Sex==MortData$Sex[i]&Age==MortData$Age[i]]$ratio
})

```

We can now create the fourth figure:

```{r}
ggplot(MortData, aes(x = ex, y = AdjustedEx, color = Age)) + geom_jitter(size = 0.1, width = 1, height = 1) +
  facet_grid(cols = vars(Sex)) + geom_abline(intercept = 0, slope = 1, color = "red") +
  labs(x = "Expected number of years of life remaining, Hungarian life table [year]",
       y = "Expected number of years of life remaining, Hanlon et al survival model [year]")
ggsave("Fig4.eps", width = 174, height = 174, units = "mm", device = cairo_ps)
```

And calculate the adjusted YLLs:

```{r}
sum(MortData$AdjustedEx) # YLL
sum(MortData$AdjustedEx) / nrow(MortData) # YLL per death
tapply(MortData$AdjustedEx, MortData$Sex, mean)
sum(MortData$AdjustedEx) / (as.numeric(difftime(EndDate, StartDate, units = "days"))/365.24 * Pop50plus) # YLL per person-year
100-sum(MortData$AdjustedEx)/sum(MortData$ex)*100
```

Life-table norms:

```{r}
sum(LifeTable[Sex=="Female", .(dx*ex)]$V1)/sum(LifeTable[Sex=="Female"]$dx)
sum(LifeTable[Sex=="Male", .(dx*ex)]$V1)/sum(LifeTable[Sex=="Male"]$dx)
sum(LifeTable[, .(dx*ex)]$V1)/sum(LifeTable$dx)
```
